n---
title: 'Lab0: Initial Setup'
author: "Stats215: Stochastic Models for Biology"
date: "1/9/2024"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#generate a transition matrix with 4 states A, C, G, T
#each state has a 25% chance of transitioning to any other state
#the transition matrix is a 4x4 matrix with each row summing to 1
#the transition matrix is a stochastic matrix
#the transition matrix is a Markov matrix

## Load libraries
```{r}
library(igraph)
```

```{r}
# generate random transition matrix 4x4 with each row summing to 1
transition_matrix <- matrix(runif(16), nrow=4, ncol=4)
transition_matrix <- transition_matrix/rowSums(transition_matrix)
#convert transition matrix to table with columns A , C, G, T
colnames(transition_matrix) <- c("A", "C", "G", "T")
rownames(transition_matrix) <- c("A", "C", "G", "T")
#plot a graph of transition matrix above with directed edges
g <- graph.adjacency(transition_matrix, mode="directed", weighted=TRUE)
plot(g, edge.label=round(E(g)$weight, 2), vertex.size=20, layout=layout.fruchterman.reingold)

```

